---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
1. Load libraries and data
```{r, echo=TRUE, tidy=TRUE}
library(lattice)
activity <- read.csv("activity.csv")
```
2. Look at structure of loaded data
```{r, echo=TRUE, tidy=TRUE}
str(activity)
head(activity)
```
3. Preprocess data as required for next steps (create subset with NA values removed)
```{r, echo=TRUE, tidy=TRUE}
act.no.na <- activity[!is.na(activity),]
```

## What is mean total number of steps taken per day?
1. Generate a histogram of the total number of steps taken each day.
```{r, echo=TRUE, tidy=TRUE}
act.date.sum <- aggregate(act.no.na$steps, by=list(date=act.no.na$date), FUN=sum)
hist(act.date.sum$x, xlab = "Total number of steps taken each day", 
        main = "Total number of steps taken each day")
```
2. Generate Mean and Median number of steps taken each day.
```{r, echo=TRUE, tidy=TRUE}
act.date.mean <- aggregate(act.no.na$steps, by=list(date=act.no.na$date), FUN=mean)
act.date.med <- aggregate(act.no.na$steps, by=list(date=act.no.na$date), FUN=median)
act.date.summary <- cbind(act.date.sum, act.date.mean$x, act.date.med$x)
names(act.date.summary) <- c("date", "sum", "mean", "median")
rm(act.date.sum, act.date.mean, act.date.med)
act.date.summary
```

## What is the average daily activity pattern?
1. Make a time series plot of the 5-min interval (x-axis) and the average number of steps taken, averaged across all the days (y-axis).
```{r, echo=TRUE, tidy=TRUE}
act.int.mean <- aggregate(act.no.na$steps, by=list(interval=act.no.na$interval), FUN=mean)
names(act.int.mean) <- c("interval", "mean")
plot(act.int.mean$interval, act.int.mean$mean, type="l",
        xlab = "Time Interval", ylab = "Mean # of Steps",
        main = "Mean Number of Steps Recorded by Time Interval")
```
2. Which 5-minute interval, on average across all the days in the data set, contains the maximum number of steps?
```{r, echo=TRUE, tidy=TRUE}
act.int.mean[which.max(act.int.mean[,2]),]
```
Answer: 835

## Imputing missing values
1. Calculate and report the total number of missing values in the data set.
```{r, echo=TRUE, tidy=TRUE}
sum(is.na(activity))
```
Answer: 2304
2. Devise a strategy to impute the missing values of the data set.
Decision: Fill in missing values with the mean across all days for that particular 5-minute interval.
3. Create a new dataset that is equal to the original, but with missing data imputed.
```{r, echo=TRUE, tidy=TRUE}
act.impute <- activity
for (i in 1:nrow(act.impute)) {
    if (is.na(act.impute[i,])) {
        int.mean <- act.int.mean[act.int.mean$interval == act.impute$interval[i],2]
        act.impute$steps[i] <- int.mean
    }
}
```
4. Make a histogram of the total number of steps taken each day
```{r, echo=TRUE, tidy=TRUE}
hist(act.date.sum2$x, xlab = "Total number of steps taken each day (Imputed)", 
     main = "Total number of steps taken each day")
```
5. Calculate and report the mean and median total number of steps taken per day.
```{r, echo=TRUE, tidy=TRUE}
act.date.mean2 <- aggregate(act.impute$steps, by=list(date=act.impute$date), FUN=mean)
act.date.med2 <- aggregate(act.impute$steps, by=list(date=act.impute$date), FUN=median)
act.date.summary2 <- cbind(act.date.sum2, act.date.mean2$x, act.date.med2$x)
names(act.date.summary2) <- c("date", "sum", "mean", "median")
rm(act.date.sum2, act.date.mean2, act.date.med2)
```
6. Do these values differ from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r, echo=TRUE, tidy=TRUE}
summary(act.date.summary$sum)
summary(act.date.summary2$sum)
```
> summary(act.date.summary$sum)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
     41    8841   10760   10770   13290   21190 
> summary(act.date.summary2$sum)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
     41    9819   10770   10770   12810   21190 

Answer: The means, medians, mins and max values are not strikingly different between the two populations, but the 1st and 3rd quartiles indicate a left shift in skew. (1st quartile was higher, 3rd quartile was lower)

## Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels: "weekday" and "weekend" indicating whether the given day is a weekday or weekend day.
```{r, echo=TRUE, tidy=TRUE}
dates <- as.Date(act.impute$date, format="%Y-%m-%d")
act.impute$weekday <- ifelse(
        weekdays(dates, abbreviate = TRUE) %in% c("Sat","Sun"),
        "weekend", "weekday")
```
2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r, echo=TRUE, tidy=TRUE}
act.int.mean2 <- aggregate(act.impute$steps, by=list(interval=act.impute$interval, weekday = act.impute$weekday), FUN=mean)
names(act.int.mean2) <- c("interval", "weekday", "mean")
xyplot(mean ~ interval | weekday, act.int.mean2, type="l",
     xlab = "Time Interval", ylab = "Mean # of Steps",
     main = "Mean Number of Steps Recorded by Time Interval (Imputed)")
```
